---
title: "Stat 453 Project Rough Draft"
author: "Erin Franke"
date: "3/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and data

```{r warning=FALSE}
library(Lahman)
library(XML)
library(ggplot2)
library(tidyr)
library(broom)
library(tidyverse)
library(dplyr)
library(car)
library(data.table)
library(stringr)
library(alluvial)
library (glmnet)
library(readr)
library(lubridate)
library(readxl)
library(writexl)
library(survival)
library(survminer)
```


Read in the draft data. This gives all players drafted from 1965 to 2019. Also read in the debut excel file, which gives all MLB debuts from 2005 to 2019. 
```{r}
draftdata <- read_csv("draft_1965_2019-2.csv") 
debuts <- read_xlsx("2005-2019 Debuts.xlsx")
```

## Data cleaning

Filter and clean the draft data, then join it with the debut data, keeping all players that were drafted between 2005 and 2019 but did not debut. Reorder the columns in the joined data set to make them easier to read and make appropriate columns be dates. 
```{r}
#change the draftdata birth_date column to be a date format and filter for draft years during and after 2005. Additionally, define a logical variable "hs" that is TRUE if the player was drafted out of high school. This is done by searching for "HS" in school using a regular expression. 
draftdata$birth_date <- as.Date(draftdata$birth_date)
draftdata519 <- draftdata %>%
  filter(year >= 2005 ) %>%
  mutate(hs = grepl("\\bHS\\b", school)) %>%
  #filter to only have the June Amateur draft included
  filter(draft_type == "JR")

#join the drafts data with the debut data; clean and organize data 
joined <- draftdata519 %>%
  left_join(debuts, by = c("name_first_last" = "Name", "birth_date" = "Birthdate"))

joined <- joined %>%
  select(name_first_last, name_last_first, primary_position, bats, throws, birth_date, weight, team_full, team_abbrev, league_full, school, hs, avg, hr, rbi, so, sb, w, l, era, sv,
year, overall, round, pick, round_sort, round_desc, draft_type, description,mlb_g, milb_g, Debut, `Last Game`, WAR, Birthplace)

joined$Debut <- as.Date(joined$Debut)
joined$`Last Game` <- as.Date(joined$`Last Game`)
```

In order to have a proper analysis, we must also remove player duplicates. These players were drafted multiple times - we will take their latest draft year. 
```{r}
newdata <- joined %>% 
  group_by(name_first_last, birth_date) %>%
  filter(year == max(year))
```

Recode the mlb_g and milb_g variable to make Y as 1 and N as 0. 
```{r}
#record mlb_g and milb_g
newdata$mlb_g <- as.integer(newdata$mlb_g == "Y")
newdata$milb_g <- as.integer(newdata$milb_g == "Y")
```

In order to allow for right censoring later on, replace all that have not made their debut in the mlb with a debut data of Sep 30, 2019. Create a variable called timetodraft that the represents the length of time between debut and draft date in days.  
```{r}
#the most recent debut in this data is Sep 27, 2019
newdata %>%
  arrange(desc(Debut))

#replace na in debut with a value of Sep 30, 2019
newdata$Debut <- replace_na(newdata$Debut, "2019-09-30") 

#create the time until draft variable
newdata$draftmonth <- rep(06,nrow(newdata)) 
newdata$draftday <- rep(10, nrow(newdata))

newdata$draftdate<-as.Date(with(newdata,paste(year,draftmonth,draftday,sep="-")),"%Y-%m-%d")
newdata <- newdata %>% select(-draftmonth, -draftday)

newdata = newdata %>% mutate(timetodraft = Debut - draftdate)

#change timetodraft variable to be an integer intead of a <time>
newdata$timetodraft <- as.integer(newdata$timetodraft)
```

## Build models

Fit a weibull model that models a player's length of time between draft and debut. Create one filtering for only players that debuted, and one that includes right censoring. 
```{r}
#fit and plot weibull model that uses right censoring 
(weiballplayers = survreg(Surv(timetodraft, mlb_g) ~ 1 , dist = "weibull" , data = newdata))
curve( 1-pweibull( x , shape=1/0.8565816 , scale=exp(9.774704)), xlim = c(0, 5000), ylim = c(0,1))

#fit weibull model that only includes players that debuted 
debutedplayers <- newdata %>%
  filter(mlb_g == 1)
(weibMlbOnly = survreg(Surv(timetodraft) ~ 1 , dist = "weibull" , data = debutedplayers))
curve( 1-pweibull( x , shape=1/0.4209225 , scale=exp(7.461209 )), xlim = c(0, 3000), ylim = c(0,1))
```

What if we were to only use include the HS variable in this model? 
```{r}
(weibMlbOnlyhs = survreg(Surv(timetodraft) ~ hs , dist = "weibull" , data = debutedplayers))
curve( 1-pweibull( x , shape=1/0.4144201, scale=exp(7.4055014)), xlim = c(0, 3000), ylim = c(0,1))
curve( 1-pweibull( x , shape=1/0.4144201, scale=exp(7.4055014+0.2140858)),add=TRUE, col="blue")
```

## Analyze the distribution of debut times

Find how long it took a player to reach the majors given that they reached the majors.
```{r}
# select June Amateur draftees drafted between 2005-2010, add hs variable, and keep latest draft year
draftdata05to10 <- draftdata %>%
  filter(2005 <= year & year <= 2010) %>%
  filter(draft_type == "JR") %>%
  mutate(hs = grepl("\\bHS\\b", school)) %>%
  # keep latest draft year
  group_by(name_first_last, birth_date) %>%
  filter(year == max(year))

# left join with debuts and reorder columns
draftdata05to10 <- draftdata05to10 %>%
  left_join(debuts, by = c("name_first_last" = "Name", "birth_date" = "Birthdate")) %>%
  select(name_first_last, name_last_first, primary_position, bats, throws, birth_date, weight,
         team_full, team_abbrev, league_full, school, hs, avg, hr, rbi, so, sb, w, l, era, sv,
         year, overall, round, pick, round_sort, round_desc, draft_type, description,mlb_g,
         milb_g, Debut, `Last Game`, WAR, Birthplace, Age)

# create draftdate and timetodebut variables
draftdata05to10 <- draftdata05to10 %>%
  mutate(draftdate = as.Date(paste(year, "6", "10", sep="-"), "%Y-%m-%d")) %>%
  mutate(across(where(is.POSIXct), as.Date)) %>%
  mutate(timetodebut = as.numeric(Debut - draftdate))
```

Let's look at the data and keep only the players who have a debut date.
```{r}
# mlb_g and Debut are slightly off due to name mismatches
sum(draftdata05to10$mlb_g == "Y")
sum(!is.na(draftdata05to10$Debut))

# here are players who reached the majors but are missing a debut date because of mismatches
# e.g. "John Mayberry Jr." in draftdata vs "John Mayberry" in drafts
draftdata05to10 %>%
  filter(mlb_g == "Y" & is.na(Debut))

# keep players who have a debut date
draftdata05to10 = draftdata05to10 %>% filter(!is.na(Debut))
```

Examine the distribution of debut times and pick a reasonable cutoff point.
```{r}
min(draftdata05to10$timetodebut, na.rm = TRUE)
max(draftdata05to10$timetodebut, na.rm = TRUE)
mean(draftdata05to10$timetodebut, na.rm = TRUE)
median(draftdata05to10$timetodebut, na.rm = TRUE)

# 95% quantile of debuts times is 8.12 years
quantile(draftdata05to10$timetodebut, 0.95) / 365


# timetodebut for hs and college draft picks
draftdata05to10 %>%
  ggplot(aes(x = timetodebut / 365, group = hs, fill = hs)) +
  geom_density(alpha = .5)

draftdata05to10 %>%
  ggplot(aes(x = timetodebut / 365)) +
  geom_histogram(aes(x = timetodebut / 365), binwidth = 0.5) +
  geom_density()

```

For players who did not debut, set their debut date to their draft date + 2964 days or the last game of the 2019 regular season, whichever comes first.

```{r}
# check that it works
newdata %>%
  filter(mlb_g == 0) %>%
  mutate(Debut = pmin(draftdate + 2964, as.Date("2019-09-30")))

newdata %>%
  mutate(Debut = if_else(mlb_g == 1, Debut, pmin(draftdate + 2964, as.Date("2019-09-30")))) %>%
  mutate(timetodebut = as.numeric(Debut - draftdate))

```

